### 2024-08-14 parts/starfree.tex
gigantic rewrite

In this section, we leverage the decidability results of \cref{polynomials:sec}
to decide $\NPoly$ inside $\ZPoly$ and $\NSF$ inside $\NPoly$, both under the
extra assumption of \kl{commutativity}. The hypothesis of \kl{commutativity}
reduces $\ZPoly$ to a combination of a finite number of polynomials, arranged
by counting modulo some numbers. This characterization will be proven in
\cref{decompose-polynomial:lem}, and will be the key ingredient in the decision
procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$ inside $\NPoly$. Note that
\cref{decompose-polynomial:lem} refines
\cref{n-poly-combinatorics:lem} for \kl{commutative} functions.

@@

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat_{\geq 1}$,
    and computable 
    polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $x_1, \dots, x_n \geq \omega$:
    \begin{equation*}
        f\left(a_1^{x_1} \cdots a_n ^{x_n}\right) 
        = P_{\moduloType[\omega](x_1, \dots, x_n)}
        \left(
            \floor{x_1/\omega}, \dots, \floor{x_n/\omega}
        \right)
        \quad .
    \end{equation*}
    \proofref{decompose-polynomial:lem}
\end{lemma}

@@

\begin{definition}
    \label{k-combinatorial:def}
    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
    such that
    for all
    $\alpha_0, \dots, \alpha_k \in \Sigma^*$,
    for all $u_1, \dots, u_k \in \Sigma^*$,
    there exists a polynomial $P \in \CorrectPoly$,  
    satisfying for all $x_1, \dots, x_k \geq \omega$:
    \begin{equation*}
        f
        \left(
            \alpha_0 \prod_{i = 1}^k u_i^{\omega x_i} \alpha_i
        \right)
        = 
        P(x_1, \dots, x_k) \quad .
    \end{equation*}
\end{definition}

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and 
    let $f \in \ZPoly[k]$ be a \kl{commutative} \kl{$\Rel$-polyregular function}.
    The following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}
\begin{proof}
    Let us prove that
    \cref{f-npoly-combi:item}
    implies \cref{f-combinatorial:item}. 
    We compute the $\omega$ by applying 
    \cref{n-poly-combinatorics:lem}, and then
    notice that the polynomial $P$ obtained is
    by definition a \kl{$\Nat$-rational polynomial}.

    For the converse implication, we use \cref{decompose-polynomial:lem} to
    effectively compute a polynomial representation of $f$.

    \textbf{TODO: this is because they coincide en enough points}.
    Using the
    assumption, all of these polynomials are actually \kl{$\Nat$-rational
    polynomials}. Let us now construct a new way to compute $f$, by first
    computing the \kl{$\omega$-type} of the input (which is possible in
    $\MSO$), and then computing the suitable \kl{$\Nat$-rational polynomial},
    which is possible in $\NPoly$.
\end{proof}

@@

Let us now prove that the above characterizations of \kl{commutative}
\kl{$\Nat$-polyregular functions} can be combined with the recent advances in
the study of \kl{$\Rel$-polyregular functions} \cite{CDTL23} to decide
aperiodicity. The key ingredient of this study is the use of a semantic
characterization of \kl{star-free $\Rel$-polyregular functions} among
\kl{$\Rel$-rational series} that generalizes the aperiodicity of languages to
functions by the means of polynomial behaviors (see
\cref{aperidic-ultimately-polynomial:ex}). Note that up until now, we only had
to consider polynomials with coefficients in $\Rel$. However, there are some
polynomials with coefficients in $\Rat$ that can be \kl{represented} by a
\kl{$\Nat$-polyregular function}. For instance, the polynomial $P(X) \defined
X(X+1)/2$. This is why, in the upcoming definition, we have to consider
polynomials with coefficients in $\Rat$.

\begin{definition}[Ultimately polynomial]
    \label{ultimately-polynomial:def}
    Let $\Sigma$ be a finite alphabet. 
    A function $f \colon \Sigma^* \to \Rel$
    is \intro{ultimately polynomial}
    when there exists $N_0 \in \Nat$ such that
    for all $n \in \Nat$
    and for all words $\alpha_0, w_1, \alpha_1, \cdots, \alpha_{n-1}, w_n, \alpha_n
    \in \Sigma^*$, there exists a polynomial $P \in \Rat[X_1, \dots, X_n]$
    such that:
    \begin{equation*}
        f\left(
            \alpha_0 \prod_{i = 1}^{n} w_i^{x_i} \alpha_i
        \right)
        = 
        P(x_1, \dots, x_n)
        \quad 
        \forall x_1, \dots, x_n \geq N_0
        \quad .
    \end{equation*}
\end{definition}

Let us recall that a language $L$ is \intro(language){aperiodic} whenever it is
recognized by an \kl(monoid){aperiodic} finite monoid, which turns out to be
equivalent to the following characterisation among regular languages: for all
$u,v,w$, there exists $n \in \Nat$, $u v^n w \in L$ if and only if $u v^{n+1} w
\in L$. As a consequence of this characterization, a language $L$ is aperiodic
if and only if its indicator function $\ind{L}$ is \kl{ultimately polynomial}.
This was already observed in \cite[Claim V.6]{CDTL23}, and reproven in
\cite[Claim 7.45, Lemma 7.53]{DOUE23}. We include a self-contained proof
hereafter for readability and didactic purposes.

@@

\AP The decidability of aperiodicity for \kl{$\Rel$-polyregular functions}, that is
the membership of $\ZSF$ inside $\ZPoly$,
relies on the construction of a canonical object called the \emph{residual
transducer}, the latter being essentially based on differences between
functions, crucially leveraging negative outputs \cite{CDTL23}. Although the
proof method does not carry from $\Rel$-output functions to $\Nat$-output
functions, it was conjectured that the semantic property of being
\kl{ultimately polynomial} would also characterize $\NSF$ inside $\NPoly$, a
conjecture that is restated hereafter.

>>>>

In this section, we leverage the decidability results of \cref{polynomials:sec}
to decide $\NPoly$ inside $\ZPoly$ and $\NSF$ inside $\NPoly$, both under the
extra assumption of \kl{commutativity}. The hypothesis of \kl{commutativity}
reduces $\ZPoly$ to a combination of a finite number of polynomials, arranged
by counting modulo some numbers. This characterization will be proven in
\cref{decompose-polynomial:lem}, and will be the key ingredient in the decision
procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$ inside $\NPoly$. Note that
\cref{decompose-polynomial:lem} can be understood as a refinement of
\cref{n-poly-combinatorics:lem} for \kl{commutative} functions.

@@

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat_{\geq 1}$,
    and computable 
    \kl{integer binomial polynomials} $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $x_1, \dots, x_n \geq 0$:
    \begin{equation*}
        f\left(a_1^{x_1} \cdots a_n ^{x_n}\right) 
        = P_{\moduloType[\omega](x_1, \dots, x_n)}
        \left(
            \floor{x_1/\omega}, \dots, \floor{x_n/\omega}
        \right)
        \quad .
    \end{equation*}
    \proofref{decompose-polynomial:lem}
\end{lemma}

@@

\begin{definition}
    \label{k-combinatorial:def}
    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
    such that
    for all
    \kl{polynomial functions} $q \colon \Nat^k \to \Sigma^*$,
    there exists a polynomial $P \in \CorrectPoly$,
    satisfying:
    \begin{equation*}
        f \circ q(\omega X_1,\dots, \omega X_k)
        = 
        P
        \quad 
        \text{ over } (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}
\end{definition}

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and 
    let $f \in \ZPoly[k]$ be a \kl{commutative} \kl{$\Rel$-polyregular function}.
    The following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}
\begin{proof}
    The implication
    \cref{f-npoly-combi:item}
    $\implies$ \cref{f-combinatorial:item} immediately follows
    from 
    \cref{n-poly-combinatorics:lem} and \cref{corrected-version:thm}.

    For the converse implication, we use \cref{decompose-polynomial:lem} to
    effectively compute a polynomial representation of $f$. That is, we obtain
    an $\omega_1 \in \Nat$ and \kl{integer binomial polynomials}
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega_1]^k}$. Now, because $f$ is
    \kl{$(k,\Nat)$-combinatorial}, we also have a number $\omega_2 \in \Nat$
    such that for all \kl{polynomial functions} $q$, there exists an
    \kl{integer binomial polynomial} $P$ such that $f \circ q(\omega_2 \vec{X})
    = P$ over $(\Nat_{\geq 1})^k$. By considering $\omega \defined
    \omega_1 \times \omega_2$, we can without loss of generality assume that
    $\omega_1 = \omega_2$ in the rest of this proof.

    Now, consider an \kl{$\omega$-type} $t \in \ModuloTypes[\omega]^k$. By
    definition, this provides us with a selection function $\ind{S} \colon \set{1,
    \dots, k} \to \set{0,1}$, and numbers $r_1, \dots, r_k$ between $0$ and
    $\omega^2-1$ such that for all $x_1, \dots, x_k \geq 1$, the tuple $(x_1
    \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has type $t$.
    In particular, for all $x_1, \dots, x_k \geq 1$, the following
    equality holds:
    \begin{equation*}
        f\left(\prod_{i = 1}^k 
            a_i^{ x_i \omega \ind{S}(i) + r_i}
        \right)
        =
        P_t(\seqof{x_i}{i \in S})
    \end{equation*}

    Let us therefore consider the \kl{polynomial function} $q \colon \Nat^k \to
    \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
    1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. Because $f$ is
    \kl{$(k,\Nat)$-combinatorial} (with parameter $\omega$), there exists
    a polynomial $P \in \CorrectPoly$ such that for all $x_1, \dots, x_n
    \geq 1$, the equality $f \circ q(\omega x_1, \dots, \omega x_k) = P(x_1,
    \dots, x_k)$ holds. In particular, we conclude that over $(\Nat_{\geq
    1})^{\card{S}}$ the polynomial $P_t$ and $P$ coincide. They are therefore
    equal as multivariate polynomials, and we conclude that $P_t$ belongs to
    $\CorrectPoly$, and in particular is an \kl{$\Nat$-rational polynomial}
    thanks to \cref{corrected-version:thm}.

    Let us now construct a new way to compute $f$, by first
    computing the \kl{$\omega$-type} of the input (which is possible in
    $\MSO$), and then computing the suitable \kl{$\Nat$-rational polynomial},
    which is possible in $\NPoly$.
\end{proof}

@@

Let us now prove that the above characterizations of \kl{commutative}
\kl{$\Nat$-polyregular functions} can be combined with the recent advances in
the study of \kl{$\Rel$-polyregular functions} \cite{CDTL23} to decide
aperiodicity. The key ingredient of this study is the use of a semantic
characterization of \kl{star-free $\Rel$-polyregular functions} among
\kl{$\Rel$-rational series} that generalizes the aperiodicity of languages to
functions by the means of polynomial behaviors (see
\cref{aperidic-ultimately-polynomial:ex}). Note that up until now, we only had
to consider polynomials with coefficients in $\Rel$. However, there are some
polynomials with coefficients in $\Rat$ that can be \kl{represented} by a
\kl{$\Nat$-polyregular function}. For instance, the polynomial $P(X) \defined
X(X+1)/2$. Historically, the definition used polynomials with coefficients in
$\Rat$, which is why we use such polynomials here, even though we know that all
of those polynomials will actually be \kl{integer binomial polynomials} thanks
to \cref{integer-binomial-polynomials:cor}.

\begin{definition}[Ultimately polynomial]
    \label{ultimately-polynomial:def}
    Let $\Sigma$ be a finite alphabet. 
    A function $f \colon \Sigma^* \to \Rel$
    is \intro{ultimately polynomial}
    when there exists $N_0 \in \Nat$ such that
    for all $n \in \Nat$,
    for all \kl{polynomial function} $q \colon \Nat^n \to \Sigma^*$,
    there exists a polynomial $P \in \Rat[X_1, \dots, X_n]$
    such that:
    \begin{equation*}
        f \circ p = P
        \quad 
        \text{ over } (\Nat_{\geq N_0})^n
        \quad .
    \end{equation*}
\end{definition}

Let us recall that a language $L$ is \intro(language){aperiodic} whenever it is
recognized by an \kl(monoid){aperiodic} finite monoid, which turns out to be
equivalent to the following characterization among regular languages: for all
$u,v,w$, there exists $n \in \Nat$, $u v^n w \in L$ if and only if $u v^{n+1} w
\in L$. As a consequence of this characterization, a language $L$ is aperiodic
if and only if its indicator function $\ind{L}$ is \kl{ultimately polynomial}.
This, was already observed in \cite[Claim V.6]{CDTL23}, and restated in
\cite[Claim 7.45, Lemma 7.53]{DOUE23}. We include a self-contained proof
hereafter for readability and didactic purposes. The proof follows essentially
the same scheme as the base case of \cref{n-poly-combinatorics:lem}.

@@

\AP The decidability of aperiodicity for \kl{$\Rel$-polyregular functions}, that is
the membership of $\ZSF$ inside $\ZPoly$,
relies on the construction of a canonical object called the \emph{residual
,
transducer}, the latter being essentially based on differences between
functions, crucially leveraging negative outputs \cite{CDTL23}. Although the
proof method does not carry from $\Rel$-output functions to $\Nat$-output,
functions, it was conjectured that the semantic property of being
\kl{ultimately polynomial} would also characterize $\NSF$ inside $\NPoly$, a
conjecture that is restated hereafter.
